---
title: "final"
author: "Hambee Makinoda"
date: 2026-03-04
format:
  html:
    toc: true 
execute:
  message: false 
  warning: false 
---

[GitHub Repo](https://github.com/Hambanana0923/ENVS-193DS_winter-2026_final)

Questions for An: Problem 1, part c (Do I have to rewrite both, and if the texts are right way)

# Set Up

```{r}
#| label: reading in packages/data
#| message: false
library(tidyverse) # read in tidyverse package
library(janitor) # read in janitor package
library(here) # read in here package
library(showtext) # read in showtext package

kelp <- read_csv(here("data", "kelp.csv")) # save giant kelp frond data in new object called kelp
mopl <- read_csv(here("data", "mopl_2019.csv")) # store Mountain Plover data in new object molp
```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, they used a linear regression test, because they were testing whether one continuous predictor variable, precipitation, predicts another continuous variable, flooded wetland area.\
\
In part 2, they used a Kruskal–Wallis test, because they were comparing the median flooded wetland area across more than two groups: wet, above normal, below normal, dry, and critical drought. Differences in median can best be measured by nonparametric test, which in this case is Kruskal-Wallis test instead of ANOVA test.\

## b. Figure needed

My coworker should use box and whisker plot to give context to the test. The x-axis should show water year classification, with the five categories: wet, above normal, below normal, dry, and critical drought. The y-axis should show flooded wetland area (m²). Additionally, my coworker should include line within the box for medians, edge of the box showing interquartile range, whiskers to show spread, and individual points beneath the boxplot to show distribution. This will visually show the readers how much flooded wetland areas between 5 water year classifications side by side, which fits well with a Kruskal–Wallis test because that test compares differences among groups without assuming normality (non-parametric).\

## c. Suggestions for rewriting

Part 1: Precipitation was not a significant predictor of flooded wetland area in the San Joaquin River Delta. This means that increase in precipitation for a year does not mean it increases flooded wetland area, and vice versa. (linear regression, F(df for whole model) = F statistic for full model , R^2^ = coefficient of determination for model, p = 0.11, α = significance level).\
\
Part 2: Median flooded wetland area did not differ significantly among the five water year classifications- wet, above normal, below normal, dry, and critical drought. (Kruskal–Wallis test, χ^2^(df) = test statistic, p = 0.12, α = significance level).\

## d. Interpretation

One additional variable that could influence flooded wetland area is soil type, which is a categorical variable. Wetlands with soils that drain slowly, such as clay-rich soils, can stay flooded longer and have larger flooded areas than wetlands with grainier/permeable soils that drain more quickly.

# Problem 2. Data visualization

Reed, D, R. Miller. 2025. SBC LTER: Reef: Kelp Forest Community Dynamics: Abundance and size of Giant Kelp (Macrocystis Pyrifera), ongoing since 2000 ver 30. Environmental Data Initiative. <https://doi.org/10.6073/pasta/c05388280c63f9effa0c71bf5908be1e>. Accessed 2026-03-11.

## a. Cleaning and summarizing

```{r}
#| label: cleaning and summarizing

kelp_clean <- kelp |> # store kelp data in new object kelp_clean
    clean_names() |> # clean column names so it only has lowercase and underscores
  
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # keeps only observations from Isla Vista, Naples, and Carpinteria
  
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # removes the single unwanted observation from transect 6, quad 40, on 2000-09-22
  
  mutate(site_name = case_when( # creates a new descriptive site_name variable from the site codes
    site == "NAPL" ~ "Naples", # mutate site name from NAPL to Naples
    site == "IVEE" ~ "Isla Vista", # mutate site name from IVEE to Isla Vista
    site == "CARP" ~ "Carpinteria" # mutate site name from CARP to Carpinteria
  )) |> 
  
  mutate(site_name = as_factor(site_name), # converts site_name to a factor
         site_name = fct_relevel(site_name,
                                 "Naples", "Isla Vista", "Carpinteria")) |> # set the order of levels manually
  
  group_by(site_name, year) |> # groups the data by site and year so summary statistics are calculated within each group
  
  summarize(mean_fronds = mean(fronds)) |> # calculates mean frond count for each site-year group
  ungroup() # removes grouping so the result is a regular data frame

kelp_clean |> slice_sample(n = 5) # slice 5 samples from the kelp_clean data frame 
str(kelp_clean) # display structure of kelp_clean
```

## b. Understanding the data

That observation was filtered out because the notes on the dataset table csv itself says there was a missing data sheet for 2000-09-22, transect 6 (a fixed 40 m sampling line in the kelp forest), quad 40 (one specific smaller sampling section within that transect where kelp were counted), so the fronds value there is -99999. That value does not represent a real number of fronds; it is an NA/missing-data code used when data were not collected or the datasheet was lost, as described in the Santa Barbara Coastal LTER data, Methods and Protocols section.

## c. Visualize the data

```{r}
#| label: mean kelp size vs sites

font_add_google("lato") # downloads/imports the Google font "Lato" 
showtext_auto() # render text in plots using showtext, so the custom font appears correctly

# ggplot base layer call
ggplot(data = kelp_clean, # select kelp_clean data frame
       aes(x = year, # use year column as x-axis
           y = mean_fronds, # use mean_fronds column as y-axis
           color = site_name)) + # color based on sites 
  geom_point() + # first layer: points
  geom_line() + # second layer: lines connecting points
  theme_minimal() + # choose minimal ggplot theme
  labs( # labels
    title = "Sites vary in mean kelp fronds per year", # label title
    x = "Year", # label x-axis title
    y = "Mean kelp fronds per year" # label y-axis title
  ) +
  scale_color_manual( # manually change colors
    name = "Site", # label legend as Site
    values = c(
      "Naples" = "#2A9D9A", # assign light blue color to Naples plot
      "Isla Vista" = "#C9A294", # assign peach color to Naples plot
      "Carpinteria" = "#7E9190")) + # assign gray color to Naples plot
  theme( # theme function call
    plot.title = element_text(hjust = 0.5, size = 14), # adjust title position and size
    panel.border = element_blank(), # eliminate plot borders
    text = element_text(family = "lato", color = "black"), # customize text font and color
    axis.title = element_text(size = 13), # change axis font size
    axis.text = element_text(size = 11), # change axis text size
    legend.title = element_text(size = 13), # change legend title font size
    legend.text = element_text(size = 11)) # change legend text size

```

## d. Interpretation

**Isla Vista** had the most variable mean kelp fronds per year because its line and points shows the largest fluctuations over time, ranging from around 5 to 40 fronds per year. The steep rises and drops in the Isla Vista series indicate much greater year-to-year variation than the other 2 sites.

**Carpinteria** had the least variable mean kelp fronds per year because its line stays relatively flatter and with a narrower range of y-values than other 2 sites, consistently around 3 to 13 fronds per year with 1 outlier point. Carpinteria shows smaller changes from year to year, with only 1 year that extends beyond 15 fronds compared to other 2 sites with multiple years over 15 fronds per year.

# Problem 3. Data Analysis

## a. Response variable

In this study, 1 represents a site that was used as a Mountain Plover nest site, and 0 represents an unused available site within a prairie dog colony. The Methods states that the nest-site selection analysis used a “use vs. non-use framework,” where used sites were actual nests and unused sites were randomly selected points within colonies where no plovers or nests were detected that year.

## b. Purpose of study

**Visual obstruction:** Visual obstruction is a continuous variable, measured with Robel pole readings at the nest and nearby points, so it represents how much vegetation height and density block visibility. As visual obstruction increases, the probability of Mountain Plover nest site use would likely decrease because the paper describes plovers select nest sites with low vegetation height. Relation of low vegetation height and Mountain Plover nesting was described in Introduction, paragraph 2, and. how/what value was measured for visual obstruction was found in the Methods section, Adult Mountain Plover Density, paragraph 3 where visual collection measurements were described.

**Distance to prairie dog colony edge**: Distance to prairie dog colony edge is also a continuous variable, because it measures how far a site is from the edge of a prairie dog colony. As distance from the prairie dog colony edge increases, the probability of Mountain Plover nesting would likely decrease because the prairie dog activity creates short vegetation and open ground that Mountain Plovers favor. The relation between Mountain Plover and prairie dogs is described in Abstract, and the variable is described in Methods, Nest-Site Selection, Table 1, where one of the predictor variable used in the nest-site selection model is described as "Distance to colony edge".

## c. Table of models

| Model number | Visual Obstruction | Distance to Prairie Dog Colony | Model Description |
|----------------|:--------------:|:--------------:|:----------------------:|
| 0 |  |  | no predictors (null model) |
| 1 | X | X | all predictors (full model): Visual Obstruction + Distance to Prairie Dog Colony |
| 2 | X |  | Visual Obstruction |
| 3 |  | x | Distance to Prairie Dog Colony |

## d. Run the models

```{r}
#| label: run model 

mopl_clean <- mopl |> # store mopl data frame in new object called mopl_clean
  clean_names() |> # clean column names so it only has lowercase letters and underscores
  mutate(
    used_bin = if_else(id == "used", 1, 0)) |> # create used_bin column and assign 1 to used nests, and 0 to others
  select(used_bin, edgedist, vo_nest) # select columns to save in the new object mopl_clean

# model 0: null model
model0 <- glm( # generalized linear model
  used_bin ~ 1, # no predictors 
  data = mopl_clean, # use mopl_clean data frame
  family = "binomial" # binomial response variable
)

# model 1: saturated model
model1 <- glm( # generalized linear model
  used_bin ~ vo_nest + edgedist, # all predictors
  data = mopl_clean, # use mopl_clean data frame
  family = "binomial" # binomial response variable
)

# model 2: visual obstruction only
model2 <- glm( # generalized linear model
  used_bin ~ vo_nest, # visual obstruction predictor only
  data = mopl_clean, # use mopl_clean data frame
  family = "binomial" # binomial response variable
)

# model 3: distance to prairie dog colony only
model3 <- glm( # generalized linear model
  used_bin ~ edgedist, # distance from prairie dog colony predictor only 
  data = mopl_clean, # use mopl_clean data frame
  family = "binomial" # binomial response variable
)
```

## e. Select the best model

```{r}
#| label: aic

AIC(model0, model1, model2, model3) #Akaike's Informatio Criterion code for all models 

```

The best model as determined by Akaike's Information Criterion (AIC) included visual obstruction at the nest (vo_nest) as a predictor variable of Mountain Plover nest site use (used\_\_bin).
