---
title: "final"
author: "Hambee Makinoda"
date: 2026-03-04
format:
  html:
    toc: true 
execute:
  message: false 
  warning: false 
---

[GitHub Repo](https://github.com/Hambanana0923/ENVS-193DS_winter-2026_final)

Questions for An: Problem 1, part c (Do I have to rewrite both, and if the texts are right way)

Problem 2, part c (legend position)

# Set Up

```{r}
#| label: reading in packages/data
#| message: false
library(tidyverse) # read in tidyverse package
library(janitor) # read in janitor package
library(here) # read in here package
library(showtext) # read in showtext package

kelp <- read_csv(here("data", "kelp.csv")) # save giant kelp frond data in new object called kelp
```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, they used a linear regression test, because they were testing whether one continuous predictor variable, precipitation, predicts another continuous variable, flooded wetland area.\
\
In part 2, they used a Kruskal–Wallis test, because they were comparing the median flooded wetland area across more than two groups: wet, above normal, below normal, dry, and critical drought. Differences in median can best be measured by nonparametric test, which in this case is Kruskal-Wallis test instead of ANOVA test.\

## b. Figure needed

My coworker should use box and whisker plot to give context to the test. The x-axis should show water year classification, with the five categories: wet, above normal, below normal, dry, and critical drought. The y-axis should show flooded wetland area (m²). Additionally, my coworker should include line within the box for medians, edge of the box showing interquartile range, whiskers to show spread, and individual points beneath the boxplot to show distribution. This will visually show the readers how much flooded wetland areas between 5 water year classifications side by side, which fits well with a Kruskal–Wallis test because that test compares differences among groups without assuming normality (non-parametric).\

## c. Suggestions for rewriting 

Part 1: Precipitation was not a significant predictor of flooded wetland area in the San Joaquin River Delta. (linear regression, F(df for whole model) = F statistic for full model , R^2^ = coefficient of determination for model, p = 0.11, α = significance level).\
\
Part 2: Median flooded wetland area did not differ significantly among the five water year classifications- wet, above normal, below normal, dry, and critical drought. (Kruskal–Wallis test, χ^2^(df) = test statistic, p = 0.12, α = significance level).\

## d. Interpretation

One additional variable that could influence flooded wetland area is soil type, which is a categorical variable. Wetlands with soils that drain slowly, such as clay-rich soils, can stay flooded longer and have larger flooded areas than wetlands with grainier/permeable soils that drain more quickly.

# Problem 2. Data visualization

Reed, D, R. Miller. 2025. SBC LTER: Reef: Kelp Forest Community Dynamics: Abundance and size of Giant Kelp (Macrocystis Pyrifera), ongoing since 2000 ver 30. Environmental Data Initiative. <https://doi.org/10.6073/pasta/c05388280c63f9effa0c71bf5908be1e>. Accessed 2026-03-11.

## a. Cleaning and summarizing 

```{r}
#| label: cleaning and summarizing

kelp_clean <- kelp |> # store kelp data in new object kelp_clean
    clean_names() |> # clean column names so it only has lowercase and underscores
  
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # keeps only observations from Isla Vista, Naples, and Carpinteria
  
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # removes the single unwanted observation from transect 6, quad 40, on 2000-09-22
  
  mutate(site_name = case_when( # creates a new descriptive site_name variable from the site codes
    site == "NAPL" ~ "Naples", # mutate site name from NAPL to Naples
    site == "IVEE" ~ "Isla Vista", # mutate site name from IVEE to Isla Vista
    site == "CARP" ~ "Carpinteria" # mutate site name from CARP to Carpinteria
  )) |> 
  
  mutate(site_name = as_factor(site_name), # converts site_name to a factor
         site_name = fct_relevel(site_name,
                                 "Naples", "Isla Vista", "Carpinteria")) |> # set the order of levels manually
  
  group_by(site_name, year) |> # groups the data by site and year so summary statistics are calculated within each group
  
  summarize(mean_fronds = mean(fronds)) |> # calculates mean frond count for each site-year group
  ungroup() # removes grouping so the result is a regular data frame

kelp_clean |> slice_sample(n = 5) # slice 5 samples from the kelp_clean data frame 
str(kelp_clean) # display structure of kelp_clean
```

## b. Understanding the data

That observation was filtered out because the notes on the dataset table csv itself says there was a missing data sheet for 2000-09-22, transect 6 (a fixed 40 m sampling line in the kelp forest), quad 40 (one specific smaller sampling section within that transect where kelp were counted), so the fronds value there is -99999. That value does not represent a real number of fronds; it is an NA/missing-data code used when data were not collected or the datasheet was lost, as described in the Santa Barbara Coastal LTER data, Methods and Protocols section.

## c. Visualize the data

```{r}
#| label: mean kelp size vs sites

font_add_google("lato") # downloads/imports the Google font "Lato" 
showtext_auto() # render text in plots using showtext, so the custom font appears correctly

# ggplot base layer call
ggplot(data = kelp_clean, # select kelp_clean data frame
       aes(x = year, # use year column as x-axis
           y = mean_fronds, # use mean_fronds column as y-axis
           color = site_name)) + # color based on sites 
  geom_point() + # first layer: points
  geom_line() + # second layer: lines connecting points
  theme_minimal() + # choose minimal ggplot theme
  labs( # labels
    title = "Sites vary in mean kelp fronds per year", # label title
    x = "Year", # label x-axis title
    y = "Mean kelp fronds per year" # label y-axis title
  ) +
  scale_color_manual( # manually change colors
    name = "Site", # label legend as Site
    values = c(
      "Naples" = "#2A9D9A", # assign light blue color to Naples plot
      "Isla Vista" = "#C9A294", # assign peach color to Naples plot
      "Carpinteria" = "#7E9190")) + # assign gray color to Naples plot
  theme( # theme function call
    plot.title = element_text(hjust = 0.5, size = 14), # adjust title position and size
    panel.border = element_blank(), # eliminate plot borders
    text = element_text(family = "lato", color = "black"), # customize text font and color
    legend.position = "right", # change legend position
    axis.title = element_text(size = 13), # change axis font size
    axis.text = element_text(size = 11), # change axis text size
    legend.title = element_text(size = 13), # change legend title font size
    legend.text = element_text(size = 11)) # change legend text size

```
